An investigation into using scagnostics to find interesting projections of high-dimensional data

### 2018-10-02 scagnostics.Rmd
Turned off caching, instead manually save results with long calculation time.

```{r initial, echo = FALSE, cache = TRUE, include = FALSE}
library(knitr)
opts_chunk$set(
  warning = FALSE, message = FALSE, echo = FALSE, 
  fig.path = 'figure/', cache.path = 'cache/', fig.align = 'center', 
  fig.show = 'hold', cache = FALSE, external = TRUE, dev = "pdf",
  fig.height = 5, fig.width = 8, out.width = "\\textwidth"
)
```

@@

[REF] (Katrin's thesis) explored the behaviour of scagnostics for selecting variables, and proposed two more, based on smoothing splines that have some nicer properties. In addition the availability of another two indices based on information criteria, maximal and total information coefficient (MIC and TIC) [@Reshef1518] round out our collection of metrics. These indices are all available in R packages, scagnostics [@HWscagR], mbgraphic [@mbgraphic] and minerva [@minerva]. The projection pursuit guided tour is available in the R package, tourr [@tourr].

@@

```{r datasets}
sphere100 <- sphereData(6, 100) %>% rescale() %>% as_tibble()
sphere1000 <- sphereData(6, 1000) %>% rescale() %>% as_tibble()

@@

```{r grandtourpath}
set.seed(1988)
t1 <- save_history(sphere100, grand_tour(2), max=4)
t1full <- as.list(interpolate(t1))
fullRes <- getProj(sin100, t1full, "sin", 100) %>%
  rbind(getProj(sin1000, t1full, "sin", 1000)) %>%
  rbind(getProj(sphere100, t1full, "sphere", 100)) %>%
  rbind(getProj(sphere1000, t1full, "sphere", 1000)) %>%
  rbind(getProj(pipe100, t1full, "pipe", 100)) %>%
  rbind(getProj(pipe1000, t1full, "pipe", 1000))
```

@@

```{r t58matrix, echo=FALSE}
knitr::kable(t1full[[58]] ,  caption = "Matrix representation of projection 58.")
```

@@

```{r plot58, fig.width=6, fig.height=7, fig.cap="View of all datasets defined by at tour step 58."}
p58 <- t1full[[58]]
pL <- list()
i <- 1
pLabels <- c("Sphere 100", "Sphere 1000", "Pipe 100", "Pipe 1000", "Sine 100", "Sine 1000")
for(ds in list(sphere100, sphere1000, pipe100, pipe1000, sin100, sin1000)){
  dm <- rescale(ds)
  pData <- as.tibble(dm %*% p58)
  pC <- ggplot(pData, aes(V1, V2)) +
    geom_point() +
    ggtitle(pLabels[i])
  pL[[i]] <- pC
  i <- i+1
}
grid.arrange(pL[[1]], pL[[2]],
             pL[[3]], pL[[4]],
             pL[[5]], pL[[6]],
             ncol=2)

@@

## Transition
We repeat the same exercise, but now we are interested in how the index values change when going from the view in V1 vs V2 to the view in V5 vs V6.
```{r plannedtourpath}
m1 <- matrix(c(1,0,0,0,0,0,0,1,0,0,0,0),ncol=2)
m2 <- matrix(c(0,0,0,0,1,0,0,0,0,0,0,1),ncol=2)
#this is silly but seems that first two entries are being ignored, so need some fake entries
m3 <- matrix(c(1,0,0,0,1,0,0,1,0,0,0,1),ncol=2)
m4 <- matrix(c(1,1,1,0,0,0,0,1,1,0,0,0),ncol=2)
t2 <- save_history(sin100,tour_path=planned_tour(list(m3,m4,m1,m2)))
t2full <- as.list(interpolate(t2))
fullResPlanned <- getProj(sin100, t2full, "sin", 100) %>%
  rbind(getProj(sin1000, t2full, "sin", 1000)) %>%
  rbind(getProj(sphere100, t2full, "sphere", 100)) %>%
  rbind(getProj(sphere1000, t2full, "sphere", 1000)) %>%
  rbind(getProj(pipe100, t2full, "pipe", 100)) %>%
  rbind(getProj(pipe1000, t2full, "pipe", 1000))
```

```{r plannedgrandtour, fig.width=6, fig.height=7, fig.cap="Short tour paths with scagnostics computed on projections. Most of the indices exhibit sharp jumps in scagnostic value."}
fullResMeltPlanned <- melt(fullResPlanned, id=c("t", "name", "size"))
ggplot(fullResMeltPlanned, aes(x=t, y=value)) +
  geom_line(aes(color=factor(size))) + 
  facet_wrap(variable~name, ncol=3, scales = "free_y", labeller = label_wrap_gen(multi_line=FALSE)) +
  guides(color=FALSE)
```

@@

```{r findpipe, results="hide"}
set.seed(2018)
pipeResc <- rescale(pipe1000) #unselect 2 variables to make optimisation easier
pipeTour <- save_history(pipeResc,
                         guided_tour(mineIndex("TIC"), search_f = tourr:::search_better,
                                     cooling=1, alpha = 0.5, max.tries = 5000))
pipeTourFull <- as.list(interpolate(pipeTour))
```

@@

```{r refine, results="hide"}
pipeTour2 <- save_history(pipeResc, guided_tour(mineIndex("TIC")), start = fProj)
pipeTourFull2 <- as.list(interpolate(pipeTour2))
pipeTourRes2 <- getProj(pipe1000, pipeTourFull2, "Pipe", 1000)
pipeTourResMelt2 <- melt(pipeTourRes2, id=c("t", "name", "size"))
ggplot(pipeTourResMelt2, aes(x=t, y=value)) +
  geom_line() + 
  facet_wrap(~variable, ncol=1, scales = "free_y")

@@

```{r findsine, fig.width=6, fig.height=7, fig.cap="Guided tour optimising the splines2d index for the Sine 1000 dataset.", results="hide", cache=TRUE}
set.seed(2018)
sineResc <- rescale(sin1000) #unselect 2 variables to make optimisation easier
sineTour <- save_history(sineResc,
                         guided_tour(splineIndex()))
sineTourFull <- as.list(interpolate(sineTour))
sineTourRes <- getProj(sin1000, sineTourFull, "Sine", 1000)
sineTourResMelt <- melt(sineTourRes, id=c("t", "name", "size"))
ggplot(sineTourResMelt, aes(x=t, y=value)) +
  geom_line() + 
  facet_wrap(~variable, ncol=1, scales = "free_y")
```

@@

```{r neutronStarEq,  results="hide"}
set.seed(2018)
nsD <- read_csv("data/samples.csv") 
nsM <- nsD %>%
  select(-m2) %>% # remove m2 to avoid finding well known correlation
  rescale() # rescaling, this returns a matrix
nseTour <- save_history(nsM, guided_tour(splineIndex()))
nseTourFull <- as.list(interpolate(nseTour))
iLast <- length(nseTourFull)
fProj <- nseTourFull[[iLast]]
dProj <- as.tibble(nsM %*% fProj)
ggplot(dProj, aes(V1,V2)) + geom_point() + theme(aspect.ratio=1)
```

>>>>

```{r initial, echo = FALSE, include = FALSE}
library(knitr)
opts_chunk$set(
  warning = FALSE, message = FALSE, echo = FALSE, 
  fig.path = 'figure/', cache.path = 'cache/', fig.align = 'center', 
  fig.show = 'hold', cache = FALSE, external = TRUE, dev = "pdf",
  fig.height = 5, fig.width = 8, out.width = "\\textwidth"
)
```

@@

[@Grimm2018] explored the behaviour of scagnostics for selecting variables, and proposed two more, based on smoothing splines that have some nicer properties. In addition the availability of another two indices based on information criteria, maximal and total information coefficient (MIC and TIC) [@Reshef1518] round out our collection of metrics. \footnote{It is interesting to note that entropy, closely related to the notion of mutual information, has previously been considered as a projection pursuit index when searching for interesting one-dimensional projections, see \cite{nason1992}, \cite{NIPS1997_1408}.}
These indices are all available in R packages, scagnostics [@HWscagR], mbgraphic [@mbgraphic] and minerva [@minerva]. The projection pursuit guided tour is available in the R package, tourr [@tourr].

@@

```{r datasets}
set.seed(1984)
sphere100 <- sphereData(6, 100) %>% rescale() %>% as_tibble()
sphere1000 <- sphereData(6, 1000) %>% rescale() %>% as_tibble()

@@

```{r grandtourpath}
set.seed(1988)
if(!file.exists("cache/grandtourpath.rda")){
  grandTourHist <- save_history(sphere100, grand_tour(2), max=4)
  grandTourInt <- as.list(interpolate(grandTourHist))
  fullRes <- getProj(sin100, grandTourInt, "sin", 100) %>%
    rbind(getProj(sin1000, grandTourInt, "sin", 1000)) %>%
    rbind(getProj(sphere100, grandTourInt, "sphere", 100)) %>%
    rbind(getProj(sphere1000, grandTourInt, "sphere", 1000)) %>%
    rbind(getProj(pipe100, grandTourInt, "pipe", 100)) %>%
    rbind(getProj(pipe1000, grandTourInt, "pipe", 1000))
  save(grandTourHist, grandTourInt, fullRes, file = "cache/grandtourpath.rda")
} else {
  load("cache/grandtourpath.rda")
}
```

@@

```{r t58matrix, echo=FALSE}
knitr::kable(grandTourInt[[58]] ,  caption = "Matrix representation of projection 58.")
```

@@

```{r plot58, fig.width=6, fig.height=7, fig.cap="View of all datasets defined by at tour step 58."}
p58 <- grandTourInt[[58]]
pL <- list()
i <- 1
pLabels <- c("Sphere 100", "Sphere 1000", "Pipe 100", "Pipe 1000", "Sine 100", "Sine 1000")
for(ds in list(sphere100, sphere1000, pipe100, pipe1000, sin100, sin1000)){
  dm <- rescale(ds)
  pData <- as.tibble(dm %*% p58)
  pC <- ggplot(pData, aes(V1, V2)) +
    geom_point() +
    ggtitle(pLabels[i])
  pL[[i]] <- pC
  i <- i+1
}
grid.arrange(pL[[1]], pL[[2]],
             pL[[3]], pL[[4]],
             pL[[5]], pL[[6]],
             ncol=2)

@@

## Transition
We repeat the same exercise, but now we are interested in how the index values change when going from the view in V1 vs V2 to the view in V5 vs V6.
```{r plannedtourpath}
if(!file.exists("cache/plannedtourpath.rda")){
  m1 <- matrix(c(1,0,0,0,0,0,0,1,0,0,0,0),ncol=2)
  m2 <- matrix(c(0,0,0,0,1,0,0,0,0,0,0,1),ncol=2)
  #this is silly but seems that first two entries are being ignored, so need some fake entries
  m3 <- matrix(c(1,0,0,0,1,0,0,1,0,0,0,1),ncol=2)
  m4 <- matrix(c(1,1,1,0,0,0,0,1,1,0,0,0),ncol=2)
  t2 <- save_history(sin100,tour_path=planned_tour(list(m3,m4,m1,m2)))
  t2full <- as.list(interpolate(t2))
  fullResPlanned <- getProj(sin100, t2full, "sin", 100) %>%
    rbind(getProj(sin1000, t2full, "sin", 1000)) %>%
    rbind(getProj(sphere100, t2full, "sphere", 100)) %>%
    rbind(getProj(sphere1000, t2full, "sphere", 1000)) %>%
    rbind(getProj(pipe100, t2full, "pipe", 100)) %>%
    rbind(getProj(pipe1000, t2full, "pipe", 1000))
  save(t2, t2full, fullResPlanned, file = "cache/plannedtourpath.rda")
} else {
  load("cache/plannedtourpath.rda")
}
```

```{r plannedtour, fig.width=6, fig.height=7, fig.cap="Short tour paths with scagnostics computed on projections. Most of the indices exhibit sharp jumps in scagnostic value."}
fullResMeltPlanned <- melt(fullResPlanned, id=c("t", "name", "size"))
ggplot(fullResMeltPlanned, aes(x=t, y=value)) +
  geom_line(aes(color=factor(size))) + 
  facet_wrap(variable~name, ncol=3, scales = "free_y", labeller = label_wrap_gen(multi_line=FALSE)) +
  guides(color=FALSE)
```

@@

```{r findpipe, results="hide"}
if(!file.exists("cache/findpipe.rda")){
  set.seed(1995)
  pipeResc <- rescale(pipe1000) 
  pipeTour <- save_history(pipeResc,
                          guided_tour(mineIndex("TIC"), search_f = tourr:::search_better,
                                     cooling=1, alpha = 0.5, max.tries = 5000))
  pipeTourFull <- as.list(interpolate(pipeTour))
  save(pipeResc, pipeTour, pipeTourFull, file = "cache/findpipe.rda")
} else {
  load("cache/findpipe.rda")
}
```

@@

```{r refinepipe, results="hide"}
if(!file.exists("cache/refinepipe.rda")){
  pipeTour2 <- save_history(pipeResc, guided_tour(mineIndex("TIC")), start = fProj)
  pipeTourFull2 <- as.list(interpolate(pipeTour2))
  pipeTourRes2 <- getProj(pipe1000, pipeTourFull2, "Pipe", 1000)
  pipeTourResMelt2 <- melt(pipeTourRes2, id=c("t", "name", "size"))
  save(pipeTour2, pipeTourFull2, pipeTourRes2, pipeTourResMelt2, file = "cache/refinepipe.rda")
} else {
  load("cache/refinepipe.rda")
}
ggplot(pipeTourResMelt2, aes(x=t, y=value)) +
  geom_line() + 
  facet_wrap(~variable, ncol=1, scales = "free_y")

@@

```{r findsine, fig.width=6, fig.height=7, fig.cap="Guided tour optimising the splines2d index for the Sine 1000 dataset.", results="hide"}
set.seed(2018)
if(!file.exists("cache/findsine.rda")){
  sineResc <- rescale(sin1000)
  sineTour <- save_history(sineResc,
                          guided_tour(splineIndex()))
  sineTourFull <- as.list(interpolate(sineTour))
  sineTourRes <- getProj(sin1000, sineTourFull, "Sine", 1000)
  sineTourResMelt <- melt(sineTourRes, id=c("t", "name", "size"))
  save(sineResc, sineTour, sineTourFull, sineTourRes, sineTourResMelt, file = "cache/findsine.rda")
} else {
  load("cache/findsine.rda")
}
ggplot(sineTourResMelt, aes(x=t, y=value)) +
  geom_line() + 
  facet_wrap(~variable, ncol=1, scales = "free_y")
```

@@

```{r neutronStarEq,  results="hide"}
set.seed(2018)
nsD <- read_csv("data/samples.csv") 
nsM <- nsD %>%
  select(-m2) %>% # remove m2 to avoid finding well known correlation
  rescale() # rescaling, this returns a matrix
if(!file.exists("cache/neutronStarEq.rda")){
  nseTour <- save_history(nsM, guided_tour(splineIndex()))
  nseTourFull <- as.list(interpolate(nseTour))
  save(nseTour, nseTourFull, file = "cache/neutronStarEq.rda")
} else {
  load("cache/neutronStarEq.rda")
}
iLast <- length(nseTourFull)
fProj <- nseTourFull[[iLast]]
dProj <- as.tibble(nsM %*% fProj)
ggplot(dProj, aes(V1,V2)) + geom_point() + theme(aspect.ratio=1)
```
